Modelling BCG vaccination in the UK

### 2019-08-01 rmds/09-model-fitting.Rmd
updated model chapter drafts

I have implemented this as a two step process for each fitted scenario. Firstly, I used the Nelder-Mead simplex method, via LibBi,[@Murray2015] to find a parameter set that optimised the maximum likelihood. I then initialised a 1000 step PMCMC chain with this parameter set, using 256 particles in the particle filter. I then used `rbi.helpers`,[@Funk:2019uw] as outlined above, to estimate the number of particles required to produce a log-likelihood variance of less than 1 for this sample of the posterior distribution, using 250 samples per step and starting with 16 particles. I initially planned to repeat this process for multiple draws from the posterior distribution but this proved to be in-feasible given the compute available. A target of 5 for the log-likelihood variance was chosen as a smaller target could not be feasibly achieved given the compute resources available. Additionally 256 was specified as the maximum number of feasible particles to use in the particle filter.

@@

In this model fitting pipeline I have used a maximum of 5 iterations of, manual, independent proposal adaption, drawing 250 samples in each iteration, starting with Gaussian distributions for each parameter, truncated by the range of the prior, with the mean based on the current parameter value. The standard deviation for each parameter was assumed to be the standard deviation of the prior if it was Gaussian and otherwise assumed to be the range of the prior if it was uniform. For each iteration I halved the  size of the standard deviation of each parameter. As for the particle calibration, I initially used a maximum likelihood method to provide a point estimate of the best fitting parameter set, followed by 1000 PMCMC steps, using a 256 particle filter. This means that the proposal distribution is adapted near to the posterior mean rather than in the tails of the posterior distribution.

@@

2. 1000 parameter sets were taken from the prior distribution and the model was then simulated for each one.

@@

4. 1000 PMCMC steps, with 256 particles used in the particle filter. This provides a better estimate of the mean of the posterior distribution.

5. Particle adaption via `rbi.helpers` at the approximate mean of the posterior distribution.[@Funk:2019uw] A minimum of 64 particles and a maximum of 256 particles are assessed with the target of a log-likelihood variance of less than 5. 250 PMCMC steps were used at each stage to estimate the log-likelihood variance.

@@

7. SMC-SMC model fitting with 1000 initial parameter particles. Particle rejuvenation was set to trigger when the effective sample size decreased below 25%, with 10 MCMC steps used each time.

@@

* Increasing the number of particles used in the outer SMC loop at the expense of reducing the number of particles used for marginal likelihood estimation.
* Increasing the number of particles used for marginal likelihood estimation. 
* Sequentially decreasing model complexity by fixing fitted parameters to manually tuned point estimates.
* Increasing the number of parameters fitted to rather than used as fixed distributions.
* Varying the  size of the proposal distribution, rejuvenation threshold, number of particles and rejuvenation steps.
* Varying the fitted observed data. This took 3 main forms:
    * Aggregation: fitting to overall incidence only; fitting to incidence grouped by large age groups (i.e children, adults, older adults);  fitting to only age groups of interest (i.e children). 
    * Reducing the time-span of the fitted data. This included simplifying down to a single year of data but also included using various combinations of time points.
    *  Exploring additional observed data. This included attempting to fit to observed pulmonary TB case from 1980 on-wards. This approach sought to constrain the parameter space to give a more realistic age distribution of cases. 
* Changing the functional form for the decay in the historic contact rate. This involved exploring linear decay with the decay gradient dictated by the year that the current contact rate takes effect and the year that the historic contact rate began to decay.
* Exploring using time dependent modifiers on the transmission probability and non-UK born mixing. 
* Exploring using modifiers for children, adults, and older adults for both the transmission probability and non-UK born mixing. This was essentially an extension of the original scenarios considered using the model fitting pipeline. 
* Exploring widening and narrowing the prior distributions of fitted parameters beyond realistic ranges. 
* Exploring varying the size of the initial high, and low, risk latent populations. This included starting with no latent cases, starting with a reduced proportion of latent cases and starting with a much larger latent population to simulate a historically more widespread disease.

@@

[@Murray2015] suggests that increasing the number of particles used in SMC may improve the quality of the model fit. Unfortunately as `LibBi` stores SMC particle paths in Random access memory (RAM) the number of particles was restricted. An additional limitation is that each rejuvenation steps also need to be stored in RAM. Attempts to increase the amount of available RAM (64 GB) using a 500GB SWAP (virtual memory) drive increased the upper limit on the number of particles but gains from this were restricted due to thrashing^[Thrashing: https://en.wikipedia.org/wiki/Thrashing_(computer_science)]. Thrashing occurs when too much data is written to SWAP memory in a short period of time and usually results in a system crash.  A major cause of the high RAM requirements of the model was that `LibBi` stores all parameters defined as initial conditions across every time point in the model, regardless of the settings used. Attempting to reduce the RAM footprint by fitting a greater number of parameters allowed for a greater number of particles to be used but also increased the model degrees of freedom and hence upped the required number of particles required by a greater amount than provided by fitting the parameter. Reducing the number of particles used to estimate the marginal likelihood allowed a greater number of particles to be used in the outer SMC step but this resulted in highly inaccurate estimates of the likelihood. These highly inaccurate likelihood estimates resulted in the SMC-SMC algorithm focusing on parameter sets that had a low likelihood estimate yet fit the data poorly - ultimately leading to even poorer model fits.

@@

After development of the model fitting pipeline but before results could be produced both the `optimise` (from `rbi`) and `adapt_particle` (from `rbi.helpers`) began to error with a difficult to debug `LibBi` error message. This meant that steps 3-5 of the model fitting pipeline could not be used with the final model. As a work around the maximum permitted number of particles (256) was used for all model fitting (increasing the number of particles was also explored as detailed in the previous section).

@@

Whilst none of the scenarios fitted the data well scenarios that included variable transmission probability in young adults fitted the data much better than those that did not (Table \@ref(tab:09-scenarios-table)). When considered on its own allowing non-UK born mixing to vary for young adults had only a small impact on the quality of the model fit in comparison to allowing the transmission probability to vary. However, the scenario that allowed both the transmission probability and non-UK born mixing to vary fit the data much better than any other scenario considered (Table \@ref(tab:09-scenarios-table)). In the following sections only the results from this scenario will be discussed.

@@

The fitted model consistently under-predicted overall TB incidence for all years with data (Table \@ref(tab:09-table-incidence-preds)). It also failed to capture the overall trend in TB incidence rates with the forecast incidence decreasing year-on-year in comparison to the observed data which showed greater variation. Stratifying by age shows that the model also failed to captured the age distribution of TB incidence (Figure \@ref(fig:09-age-strat-incidence). Whilst the model under predicted TB incidence in all age-groups it was more accurate for older adults, implying that even if the magnitude of cases had been better predicted the distribution would still not match the observed data.

>>>>

I have implemented this as a two step process for each fitted scenario. Firstly, I used the Nelder-Mead simplex method, via LibBi,[@Murray2015] to find a parameter set that optimised the maximum likelihood. I then initialised a 1000 step PMCMC chain with this parameter set, using 1024 particles in the particle filter. I then used `rbi.helpers`,[@Funk:2019uw] as outlined above, to estimate the number of particles required to produce a log-likelihood variance of less than 1 for this sample of the posterior distribution, using 250 samples per step and starting with 16 particles. I initially planned to repeat this process for multiple draws from the posterior distribution but this proved to be in-feasible given the compute available. A target of 5 for the log-likelihood variance was chosen as a smaller target could not be feasibly achieved given the compute resources available. Additionally 1024 was specified as the maximum number of feasible particles to use in the particle filter.

@@

In this model fitting pipeline I have used a maximum of 5 iterations of, manual, independent proposal adaption, drawing 250 samples in each iteration, starting with Gaussian distributions for each parameter, truncated by the range of the prior, with the mean based on the current parameter value. The standard deviation for each parameter was assumed to be the standard deviation of the prior if it was Gaussian and otherwise assumed to be the range of the prior if it was uniform. For each iteration I halved the  size of the standard deviation of each parameter. As for the particle calibration, I initially used a maximum likelihood method to provide a point estimate of the best fitting parameter set, followed by 1000 PMCMC steps, using a 1024 particle filter. This means that the proposal distribution is adapted near to the posterior mean rather than in the tails of the posterior distribution.

@@

2. 2000 parameter sets were taken from the prior distribution and the model was then simulated for each one.

@@

4. 1000 PMCMC steps, with 1024 particles used in the particle filter. This provides a better estimate of the mean of the posterior distribution.

5. Particle adaption via `rbi.helpers` at the approximate mean of the posterior distribution.[@Funk:2019uw] A minimum of 64 particles and a maximum of 1024 particles are assessed with the target of a log-likelihood variance of less than 5. 250 PMCMC steps were used at each stage to estimate the log-likelihood variance.

@@

7. SMC-SMC model fitting with 2000 initial parameter particles. Particle rejuvenation was set to trigger when the effective sample size decreased below 25%, with 10 MCMC steps used each time.

@@

* Increasing the number of particles used in the outer SMC loop at the expense of reducing the number of particles used for marginal likelihood estimation.
* Increasing the number of particles used for marginal likelihood estimation. 
* Sequentially decreasing model complexity by fixing fitted parameters to manually tuned point estimates.
* Increasing the number of parameters fitted to rather than used as fixed distributions.
* Varying the  size of the proposal distribution, rejuvenation threshold, number of particles and rejuvenation steps.
* Varying the fitted observed data. This took 3 main forms:
    * Aggregation: fitting to overall incidence only; fitting to incidence grouped by large age groups (i.e children, adults, older adults);  fitting to only age groups of interest (i.e children). 
    * Reducing the time-span of the fitted data. This included simplifying down to a single year of data but also included using various combinations of time points.
    *  Exploring additional observed data. This included attempting to fit to observed pulmonary TB case from 1980 on-wards. This approach sought to constrain the parameter space to give a more realistic age distribution of cases. 
* Changing the functional form for the decay in the historic contact rate. This involved exploring linear decay with the decay gradient dictated by the year that the current contact rate takes effect and the year that the historic contact rate began to decay.
* Exploring using time dependent modifiers on the transmission probability and non-UK born mixing. 
* Exploring using modifiers for children, adults, and older adults for both the transmission probability and non-UK born mixing. This was essentially an extension of the original scenarios considered using the model fitting pipeline. 
* Exploring widening and narrowing the prior distributions of fitted parameters beyond realistic ranges. 
* Exploring varying the size of the initial high, and low risk latent populations. This included starting with no latent cases, starting with a reduced proportion of latent cases and starting with a much larger latent population to simulate a historically more widespread disease.

@@

[@Murray2015] suggests that increasing the number of particles used in SMC may improve the quality of the model fit. Unfortunately as `LibBi` stores SMC particle paths in Random access memory (RAM) the number of particles was restricted. An additional limitation is that the current rejuvenation step also need to be stored in RAM. Attempts to increase the amount of available RAM (64 GB) using a 500GB SWAP (virtual memory) drive increased the upper limit on the number of particles but gains from this were restricted due to thrashing^[Thrashing: https://en.wikipedia.org/wiki/Thrashing_(computer_science)]. Thrashing occurs when too much data is written to SWAP memory in a short period of time and usually results in a system crash.  A major cause of the high RAM requirements of the model was that `LibBi` stores all parameters defined as initial conditions across every time point in the model, regardless of the settings used. Attempting to reduce the RAM footprint by fitting a greater number of parameters allowed for a greater number of particles to be used but also increased the model degrees of freedom and hence upped the required number of particles required by a greater amount than provided by fitting the parameter. Reducing the number of particles used to estimate the marginal likelihood allowed a greater number of particles to be used in the outer SMC step but this resulted in highly inaccurate estimates of the likelihood. These highly inaccurate likelihood estimates resulted in the SMC-SMC algorithm focusing on parameter sets that had a low likelihood estimate yet fit the data poorly - ultimately leading to poorer model fits.

@@

After development of the model fitting pipeline but before results could be produced both the `optimise` (from `rbi`) and `adapt_particle` (from `rbi.helpers`) began to error with a difficult to debug `LibBi` error message. This meant that steps 3-5 of the model fitting pipeline could not be used with the final model. As a work around the maximum permitted number of particles (1024) was used for all model fitting (increasing the number of particles was also explored as detailed in the previous section).

@@

Whilst none of the scenarios fitted the data well scenarios that included variable transmission probability in young adults fitted the data much better than those that did not (Table \@ref(tab:09-scenarios-table)). When considered on its own, allowing non-UK born mixing to vary for young adults had only a small impact on the quality of the model fit in comparison to allowing the transmission probability to vary. However, the scenario that allowed both the transmission probability and non-UK born mixing to vary fit the data much better than any other scenario considered (Table \@ref(tab:09-scenarios-table)). In the following sections only the results from this scenario will be discussed.

@@

The fitted model consistently under-predicted overall TB incidence for all years with data (Table \@ref(tab:09-table-incidence-preds)). It also failed to capture the overall trend in TB incidence rates with the forecast incidence increasing year-on-year in comparison to the observed data which showed greater variation. Stratifying by age shows that the model also failed to captured the age distribution of TB incidence (Figure \@ref(fig:09-age-strat-incidence). Whilst the model under predicted TB incidence in all age-groups it was more accurate for older adults, implying that even if the magnitude of cases had been better predicted the distribution would still not match the observed data.

### 2019-08-01 rmds/10-model-results.Rmd
updated model chapter drafts

The estimated age-stratified, and aggregated, TB incidence rates were compared both visually and numerically from 2005 through to 2040 for each scenario. From 2005-2015 observed data were also compared to the estimates from each scenario. Multiple time horizons were evaluated as initially the impact of any policy change may be masked by the large reservoir of vaccinated individuals in the populations and because of the impact of the assumed decrease in non-UK born cases over time.

@@

Continuing with school-age vaccination resulted in the fewest number of cases regardless of the time-span considered (Table \@ref(tab:10-table-future-incidence). However, the difference between vaccination scenarios was consistently small when compared to the overall number of cases. In all scenarios TB incidence was forecast to decrease over time in line with the decreases assumed in non-UK born TB incidence. The lower bounds for each scenario were relatively comparable, with the upper bounds being being higher for both neonatal BCG vaccination and no BCG vaccination when compared to school-age BCG vaccination.

@@

As expected the neonatal vaccination resulted in a dramatic drop in TB incidence rates in 0-4 year olds and a smaller but still large reduction in 5-9 year olds (Figure \@ref(fig:10-age-scenarios-fig)). There was a light reduction in 10-15 year olds. School-age vaccination resulted in the lower incidence rates in all adult populations when compared to any other scenario, except in adults over 45 where all scenarios were comparable. Neonatal vaccination resulted in a slight decrease in TB incidence rates when compared to no vaccination in young adults.

@@

```{r 10-table-deaths-preds}
sum_deaths_scenario %>% filter(Year %% 5 == 0) %>% 
  set_names(., nm = names(.) %>% str_replace_all("BCG", "BCG (95% CI)")) %>% 
  select(-Observation) %>% 
kable(
  caption.short = "Forecast overall TB mortality for each scenario evaluated from 2005 to 2040.", 
  caption = "Forecast overall TB mortality for each scenario evaluated from 2005 to 2040. For brevity only 5 year intervals are shown.(95\\% CI): 95\\% credible interval estimated using the 2.5\\% quantile and the 97.5\\% quantile.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(latex_options = c("hold_position"))
```

@@

The results presented in this chapter generally agree with other findings from this thesis. In Chapter \@ref(direct-eff), which estimated the impact of the change in vaccination policy in those directly impacted by it, there was some evidence that changing to neonatal vaccination was associated with a small increase in incidence rates in those who were school-age. There was less evidence of a reduction in incidence rates in UK born neonates. The first of these results matches the findings presented here. However, this chapter estimated a rapid reduction in TB incidence in young children which was not seen in this previous work. Chapter \@ref(sutherland), which recreated a previously published transmission chain model estimated an initial impact from ending school-age BCG vaccination but that this impact would decline with time. The results presented here agree with this findings as long as it is assumed that non-UK born incidence will decrease over time. However, here the impact of the scheme was estimated to continue beyond the 15 year time horizon estimated in Chapter \@ref(sutherland). In Chapter \@ref(beneficial-bcg-out) I found some evidence that BCG vaccination may decrease all-cause mortality in TB cases and some evidence that indicated that this may be related to reduced TB mortality. If either of these associations were causal then they would increase the benefit of vaccination scenarios compared to no vaccination but would not alter the trade-offs between neonatal and school-age vaccination.

Further work is need to improve the fit of the model to observed data (Chapter \@ref(model-fit)). This will result in improved forecasts and more reliable results. In addition additional vaccination coverage scenarios could be considered that explored the impact of vaccinating a reduced proportion of both those at school-age and neonates. If additional data becomes available, or with the appropriate assumptions, the inclusion of the high-risk population into the model would allow the evaluation of targeted high-risk neonatal vaccination in comparison to the other scenarios considered here. The extrapolation of the trend in non-UK born cases is a limitation of this model and as such should be further explored using other assumptions such as constant non-UK born cases, incidence rates based on expert opinion and estimates based on other global modelling studies.  
The results presented here indicate that changing from a school-age BCG vaccination programme to a neonatal BCG programme leads to an overall increase in incidence rates, with increases concentrated in the young adults, and to a lesser degree, in older adults. Neonatal vaccination led to a decrease in incidence rates in children both in the short - and long term. This indicates that direct vaccination provides the best protection for children rather than indirect protection via reduced transmission. This finding is likely to be dependent on the degree of background transmission and so further modelling studies are needed in diverse settings before conclusions can be generalised. No vaccination was shown to lead to increased incidence in all age-groups when compared to school-age vaccination and in children only when compared to neonatal vaccination. The impact of any vaccination programme on older adults was small. These results are preliminary in nature as the model on which they are based fitted poorly to the observed data.

>>>>

Estimated age-stratified, and aggregated, TB incidence, and mortality were compared both visually and numerically from 2005 through to 2040 for each scenario. Multiple time horizons were evaluated as initially the impact of any policy change may be masked by the large reservoir of vaccinated individuals in the population and because of the impact of the assumed decrease in non-UK born cases over time.

@@

Continuing with school-age vaccination resulted in the fewest number of cases regardless of the time-span considered (Table \@ref(tab:10-table-future-incidence)). However, the difference between vaccination scenarios was consistently small when compared to the overall number of cases. In all scenarios TB incidence was forecast to decrease over time in line with the decreases assumed in non-UK born TB incidence. The lower bounds for each scenario were relatively comparable, with the upper bounds being being higher for both neonatal BCG vaccination and no BCG vaccination when compared to school-age BCG vaccination.

@@

As expected the neonatal vaccination resulted in a rapid decline in TB incidence in 0-4 year olds and a smaller but still large reduction in 5-9 year olds (Figure \@ref(fig:10-age-scenarios-fig)). There was a slight reduction in 10-15 year olds. School-age vaccination resulted in the lower incidence rates in all adult populations when compared to any other scenario, except in adults over 45 where all scenarios were comparable. Neonatal vaccination resulted in a slight decrease in TB incidence rates when compared to no vaccination in young adults.

@@

```{r 10-table-deaths-preds}
sum_deaths_scenario %>% filter(Year %% 5 == 0) %>% 
  set_names(., nm = names(.) %>% str_replace_all("BCG", "BCG (95% CI)")) %>% 
  select(-Observation) %>% 
kable(
  caption.short = "Forecast overall TB mortality for each scenario evaluated from 2005 to 2040.", 
  caption = "Forecast overall TB mortality for each scenario evaluated from 2005 to 2040. For brevity only 5 year intervals are shown.(95\\% CI): 95\\% credible interval estimated using the 2.5\\% quantile and the 97.5\\% quantile.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(latex_options = c("HOLD_position"))
```

@@

The results presented in this chapter generally agree with other findings from this thesis. In Chapter \@ref(direct-eff), which estimated the impact of the change in vaccination policy in those directly impacted by it, there was some evidence that changing to neonatal vaccination was associated with a small increase in incidence rates in those who were school-age. There was less evidence of a reduction in incidence rates in UK born neonates. The first of these results matches the findings presented here. However, this chapter estimated a rapid reduction in TB incidence in young children which was not seen in the previous work. Chapter \@ref(sutherland), which recreated a previously published transmission chain model estimated an initial impact from ending school-age BCG vaccination but that this impact would decline with time. The results presented here agree with this findings as long as it is assumed that non-UK born incidence will decrease over time. However, here the impact of the scheme was estimated to continue beyond the 15 year time horizon estimated in Chapter \@ref(sutherland). In Chapter \@ref(beneficial-bcg-out) I found some evidence that BCG vaccination may decrease all-cause mortality in TB cases and some evidence that indicated that this may be related to reduced TB mortality. If either of these associations were causal then they would increase the benefit of vaccination scenarios compared to no vaccination but would not alter the trade-offs between neonatal and school-age vaccination.

Further work is need to improve the fit of the model to observed data (Chapter \@ref(model-fitting)). This will result in improved forecasts and more reliable results. In addition additional vaccination coverage scenarios could be considered that explored the impact of vaccinating a reduced proportion of both those at school-age and neonates. If additional data becomes available, or with the appropriate assumptions, the inclusion of the high-risk population into the model would allow the evaluation of targeted high-risk neonatal vaccination in comparison to the other scenarios considered here. The extrapolation of the trend in non-UK born cases is a limitation of this model and as such should be further explored using other assumptions such as constant non-UK born cases, incidence rates based on expert opinion and estimates based on other global modelling studies.

The results presented here indicate that changing from a school-age BCG vaccination programme to a neonatal BCG programme lead to an overall increase in incidence rates, with increases concentrated in the young adults, and to a lesser degree, in older adults. Neonatal vaccination led to a decrease in incidence rates in children both in the short - and long term. This indicates that direct vaccination provides the best protection for children rather than indirect protection via reduced transmission. This finding is likely to be dependent on the degree of background transmission and so further modelling studies are needed in diverse settings before conclusions can be generalised. No vaccination was shown to lead to increased incidence in all age-groups when compared to school-age vaccination and in children only when compared to neonatal vaccination. The impact of any vaccination programme on older adults was small. These results are preliminary in nature as the model on which they are based fitted poorly to the observed data.
